---
title: "Spatial Challenge"
author: "Farhan Ameen"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    code_folding: hide
    self_contained: yes
    theme: spacelab
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE)

```

```{r libraries}
library(tidyverse)
library(SingleCellExperiment)
library(Statial)
library(spicyR)
library(plotly)
library(ClassifyR)
library(lisaClust)
library(ggsurvfit)
library(ggthemes)
library(DT)
library(htmlwidgets)
library(ggthemes)
library(tidySingleCellExperiment)

# Set ggplot theme
theme_set(theme_classic())

```

# Loading datasets

Dataset 1 - [Keren et al](https://pubmed.ncbi.nlm.nih.gov/30193111/)

Dataset 2 - [Ali et al](https://www.nature.com/articles/s43018-020-0026-6)

The next sections will look at Keren et al.

```{r data}
#load("./data/kerenSCE.rda")
load("./data/aliSCE.rda")


#kerenSCE
aliSCE %>% 
  colData() %>% 
  data.frame() %>% 
  names()
```

# Data cleaning

```{r cleaningData}

# Removing patients with cold tumour due to small sample size of 5
kerenSCE = kerenSCE[, kerenSCE$tumour_type != "cold"]
kerenSCE$tumour_type = droplevels(kerenSCE$tumour_type)

# Removing patients 22 and 38 because they have missing survival data.
kerenSCE = kerenSCE[, !kerenSCE$imageID %in% c("22", "38")]
kerenSCE$imageID = as.numeric(kerenSCE$imageID)


# Extracting clinical information
clinicalDf = kerenSCE |> 
  colData() |> 
  data.frame() |> 
  select(-c(x, y, CellID, cellType, cellSize, C, tumorYN, tumorCluster, Group, immuneCluster, immuneGroup, region)) |> 
  unique() |> 
  remove_rownames()

```

# Data visualisation {.tabset}

## Dimension reduction plot

```{r UMAP}
#set.seed(51773)
#kerenSCE <- scater::runUMAP(kerenSCE, exprs_values = "intensities", name = "UMAP")
set.seed(51773)
aliSCE <- scater::runUMAP(aliSCE, exprs_values = "counts", name = "UMAP")


scater::plotReducedDim(kerenSCE, dimred = "UMAP", colour_by = "cellType")
```

## Example image

```{r exampleImage}
p = kerenSCE |> 
  colData() |> 
  data.frame() |> 
  filter(imageID == "4") |> 
  ggplot(aes(x = x, y = y, col = cellType)) +
  geom_point(size = 1) +
  ggthemes::scale_colour_tableau( palette = "Tableau 20")


ggplotly(p)
```

# Calculating cell proportions {.tabset}

## Proportions

```{r cellProp}
cellProp = spicyR::getProp(
  cells = kerenSCE
)

cellProp |> 
  round(4) |> 
  DT::datatable(options = list(scrollX = TRUE))
```

## Testing across conditions

```{r testProp}
testProp = spicyR::colTest(kerenSCE,
                condition = "tumour_type",
                feature = "cellType")

testProp |> 
  DT::datatable(options = list(scrollX = TRUE))

```

```{r propBoxplot}
cellProp |> 
  rownames_to_column("imageID") |> 
  mutate(imageID = as.numeric(imageID)) |> 
  left_join(clinicalDf) |> 
  ggplot(aes(x = tumour_type, y = `Keratin+Tumour`)) +
  geom_boxplot() +
  labs(y = "Proprotion of Keretin+Tumour cells",
       x = "Tumour Type")
  
```

# Region analysis

# **Max CODE CODE CODE CODE here**

```{r region}
set.seed(51773)

aliSCE <- lisaClust(
  aliSCE,
  k = 5,
  spatialCoords = c("x", "y"),
  cellType = "cellType", #the markers
  imageID = "image_number"
)



#BPARAM = MulticoreParam(workers = 2)
```

```{r region}

regionMap(aliSCE,
          cellType = "cellType")

# regionMap(aliSCE,
#           cellType = "cellType",
#           )
# #LR, DR loco or distant relapse date, 1=yes, 0=no
# regionMap(aliSCE,
#           cellType = "cellType",
#           )
```

## Visualising regions {.tabset}

### Hatching plot

```{r viewRegion, fig.height= 8, fig.width=10}
p = hatchingPlot(
  aliSCE,
  useImages = "527",
  cellType = "cellType",
  spatialCoords = c("x", "y"),
  imageID = "image_number",
  #line.spacing = 41, # spacing of lines
  #nbp = 100 # smoothness of lines
)

p +
  ggthemes::scale_colour_tableau( palette = "Tableau 20") +
  guides(fill=guide_legend(ncol=2))

```

```{r}
regionMap(aliSCE,
          cellType = "cellType")

```

### Coloured regions

```{r}

#entire relapse
aliRelapseFactor = aliSCE %>%  
  mutate(cellType = factor(cellType, ordered = TRUE, levels = sort(unique(cellType))))

aliSCELRrelapseno = aliRelapseFactor |>
  filter(lr==0, dr==0)

regionMap(aliSCELRrelapseno,
          cellType = "cellType", type = "heatmap", cluster_rows = FALSE)

aliSCELRrelapseyes = aliRelapseFactor |>
  filter(lr==1, dr==1)

regionMap(aliSCELRrelapseyes,
          cellType = "cellType", type = "heatmap",cluster_rows = FALSE)


#compare to non spatial data, prove that spatial methods have merit

```

```{r}
aliRelapseFactor = aliSCE %>%  
  mutate(cellType = factor(cellType, ordered = TRUE, levels = sort(unique(cellType))))

#loco relapse
aliSCELocorelapseno = aliRelapseFactor |>
  filter(lr==0)

regionMap(aliSCELocorelapseno,
          cellType = "cellType", type = "heatmap", cluster_rows = FALSE)

aliSCELocorelapseyes = aliRelapseFactor |>
  filter(lr==1)

regionMap(aliSCELocorelapseyes,
          cellType = "cellType", type = "heatmap",cluster_rows = FALSE)

#no loco, no distal
#yes loco, no distal recurance
aliSCELocorelapsenoalpha = aliRelapseFactor |>
  filter(lr==0, dr==0)

regionMap(aliSCELocorelapsenoalpha,
          cellType = "cellType", type = "heatmap", cluster_rows = FALSE)

aliSCELocorelapseyesalpha = aliRelapseFactor |>
  filter(lr==1, dr==0)

regionMap(aliSCELocorelapseyesalpha,
          cellType = "cellType", type = "heatmap",cluster_rows = FALSE)

#no loco, yes distal
#yes loco, yes distal recurance
aliSCELocorelapsenobeta = aliRelapseFactor |>
  filter(lr==0, dr==1)

regionMap(aliSCELocorelapsenobeta,
          cellType = "cellType", type = "heatmap", cluster_rows = FALSE)

aliSCELocorelapseyesbeta = aliRelapseFactor |>
  filter(lr==1, dr==1)

regionMap(aliSCELocorelapseyesbeta,
          cellType = "cellType", type = "heatmap",cluster_rows = FALSE)


#distal relapse
aliSCEdistalrelapseno = aliRelapseFactor |>
  filter(dr==0)

regionMap(aliSCEdistalrelapseno,
          cellType = "cellType", type = "heatmap", cluster_rows = FALSE)

aliSCEdistalrelapseyes = aliRelapseFactor |>
  filter(dr==1)

regionMap(aliSCEdistalrelapseyes,
          cellType = "cellType", type = "heatmap",cluster_rows = FALSE)


#compare to non spatial data, prove that spatial methods have merit

```

```{r colourRegion}
aliSCE |> 
  colData() |> 
  data.frame() |> 
  filter(image_number == "50") |> 
  ggplot(aes(x = x, y = y, col = region)) +
  geom_point(size=5) +
  ggthemes::scale_colour_tableau() +
  theme_classic()

#40, 5 terrible images
#different imaging technology, smaller number of cells
```

end code????????????????????????????????????????

### Region composition

```{r}
regionMap(kerenSCE,
          cellType = "cellType")
```

# Spatial interactions

A positive value means the two cell types are localised, a negative value means they are dispersed.

```{r pairwise}
interactions = spicyR::getPairwise(
  cells = kerenSCE,
) |> 
  data.frame()

interactions |> 
  round(4) |> 
  DT::datatable( options = list(scrollX = TRUE))
```

```{r spatialImage, fig.height=8, fig.width=10}
p = kerenSCE |> 
  colData() |> 
  data.frame() |> 
  filter(imageID == "3") |> 
  ggplot(aes(x = x, y = y, col = cellType)) +
  geom_point(size = 1) +
  ggthemes::scale_colour_tableau( palette = "Tableau 20") +
  theme_classic() +
  ggtitle("Image 3")


ggplotly(p)
```

## Check which spatial interactions are associated with survival

The interaction between Keratin+Tumour cells and other immune cells are the most significant spatial relationship which contributes to patient survival.

The CoxPH coefficient is negative, which means when Keratin+Tumour and other immune cells are localising, the patient tends to live longer.

```{r colCox}
# Create survival object from data 
survivalOutcomes = Surv(clinicalDf$`Survival_days_capped.`, clinicalDf$Censored)

# Fit CoxPh models on all cell relationships.
ClassifyR::colCoxTests(interactions, survivalOutcomes) |> 
  arrange(p.value) |> 
  round(4) |> 
  DT::datatable( options = list(scrollX = TRUE))

```

```{r survCurve}
# Selecting the most significant relationship
relationship = interactions$Keratin.Tumour__other.immune

# Splitting the values by median relationship
relationship = ifelse(relationship > median(relationship), "Attraction", "Avoidance")
    
# Plotting Kaplan-Meier curve
survfit2(survivalOutcomes ~ relationship) |>
    ggsurvfit() +
    add_pvalue() +
    ggtitle("Keratin+Tumour__Other immune")
```

# Classification

```{r classification}
# Calculate proportions of regions in each image
regionProp <- spicyR::getProp(kerenSCE, 
                       feature = "region",
                       imageID = "imageID")

# Get the average expression of a marker in each cell type in each region
cellTypeRegionMeans <- Statial::getMarkerMeans(kerenSCE,
                              imageID = "imageID",
                              cellType = "cellType",
                              region = "region")

# Group all features in one vector
featureList = list(
  "Cell type proportions" = cellProp,
  "Region proportions" = regionProp,
  "Marker means" = cellTypeRegionMeans,
  "Spatial interactions" = interactions
)



# Training a CoxNet survival model 
CV = ClassifyR::crossValidate(
  measurements = featureList,
  outcome = survivalOutcomes,
  classifier = "CoxNet",
  selectionMethod = "CoxPH",
  nFolds = 5,
  nFeatures = 10,
  nRepeats = 20
)

```

```{r classificationPlot}
ClassifyR::performancePlot(CV, metric = "C-index",
                characteristicsList = list(x = "auto", fillColour = "Assay Name")) +
  theme(legend.position = "none") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme(axis.text.x = element_text(size = 7)) +
  scale_fill_tableau()
```
